Principle of least action as a software development method

### 2022-11-03 ms.tex
fix

\lstdefinestyle{lststyle}{basicstyle=\tt\footnotesize, captionpos=b, frame=single}

@@

\title{A Makefile for Developing Containerized \LaTeX\ Technical Documents}

@@

\begin{abstract}
	We propose a Makefile for developing containerized \LaTeX\ technical documents.
	The Makefile allows the author to execute the code that generates variables, tables and figures (results), which are then used during the \LaTeX\ compilation, to produce either the draft (fast) or full (slow) version of the document.
	We also present various utilities that aid in automating the results generation and improve the reproducibility of the document.
	We lastly release an open source repository of this paper that uses the Makefile.
\end{abstract}

\section{Introduction}
Developing technical documents that present both computationally generated results (variables, tables and figures) and natural text, is a central aspect of computational research.
One programming paradigm that was proposed as a solution is \textit{literate programming}~\cite{knuth1984literate}, in which the author inserts snippets of code and the output of its execution, alongside natural text.
Applications of \textit{literate programming} include Jupyter~\cite{kluyver2016jupyter} which allows publishing code, results and explanations in a executable format, PythonTeX~\cite{poore2015pythontex} which allows Python to be executed in \LaTeX\ and ActivePapers~\cite{hinsen2014activepapers} in which code can be executed using the Java Virtual Machine.

Although \textit{literate programming} provides many advantages for demonstrative projects and immediate visual feedback during experimentation, its `dual nature' constrains authors in using specifically designed text editors and/or source code formats thus leading to `app/vendor lock-in' situations.
Moreover, the overlap between natural text and results code makes \textit{literate programming} difficult to debug and version control.

We propose a Makefile for developing containerized \LaTeX\ technical documents.
By using portable technologies that withstood the test of time, we ensure the stability and portability of the proposed Makefile.
We write this paper using a repository that consists of the Makefile and choose Docker as the container technology and the Python programming language for generating the results.

For the rest of the paper we will refer to:
\begin{itemize}
	\item \textit{results code} as the scientific-oriented programming language code (\textit{main.py}) that produces the variables, tables and figures,
	\item \textit{results} as the variables, tables and figures that are generated by the \textit{results code},
	\item \textit{\LaTeX\ code} as the natural text (\textit{ms.tex}) and bibliography (\textit{ms.bib}),
	\item \textit{code} as both the \textit{results code} and \textit{\LaTeX\ code},
	\item \textit{document} as the technical document that contains the rendered \textit{results} and natural text and
	\item \textit{author} as the author of the \textit{document} that develops the \textit{code}.
\end{itemize}

\section{The Makefile}

\begin{figure}[!t]
	\centering
	\begin{tikzpicture}[%
			grow via three points={one child at (0.5,-0.7) and two children at (0.5,-0.7) and (0.5,-1.4)},
			edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
			\node[draw](document){document/}
			child{node[draw, anchor=west](makefile){Makefile}}
			child{node[draw, anchor=west, fill=red!20](tex){ms.tex}}
			child{node[draw, anchor=west, fill=red!20](bib){ms.bib}}
			child{node[draw, anchor=west, fill=blue!20](py){main.py}}
			child{node[draw, anchor=west, fill=blue!20](containerfile){Containerfile}}
			child{node[draw, anchor=west, fill=blue!20](requirements){requirements.txt}};
			\begin{scope}[on background layer]
				\node[draw, rounded corners, dashed, fill=gray!20, inner sep=4pt, fit=(document) (requirements)]{};
				\node[draw, rounded corners, dashed, fill=gray!80, inner sep=2pt, fit=(py) (requirements)](code){};
				\node[draw, rounded corners, dashed, fill=gray!80, inner sep=2pt, fit=(tex) (bib)](document){};
			\end{scope}
			\node[draw, right=1.5cm of code, fill=green!20] (results) {python/bin/results};
			\path[draw, ->] (code) -- (results);
			\node[draw, right=4cm of document, fill=green!20] (pdf) {python/bin/ms.pdf};
			\path[draw, ->] (results) |- (pdf);
			\path[draw, ->] (document) -- (pdf);
	\end{tikzpicture}
	\caption{The files of a repository that uses the Makefile is depicted in the light gray background and the arrows denote the containerized data flow towards generating the \textit{document}.
	Green indicates temporary cache or generated files, red \textit{\LaTeX\ code} files, blue \textit{results code} files and within the dark gray background, files that are developed by the \textit{author}.}\label{fig:repositoryfiles}
\end{figure}

The proposed Makefile can be used for developing \textit{documents} that consist of \textit{results} generated from \textit{results code} executed within a container.
The three tools that are used by the Makefile are the Make program which is a build automation tool, \LaTeX\ which is a typesetting system for publishing high quality research documents, and the container technology (such as Docker) which allows portable execution of \textit{code} between different hardware, operating systems and environments.

The pseudoalgorithm of the Makefile is as follows:
\begin{lstlisting}[language=bash, style=lststyle, caption={Pseudoalgorithm of the Makefile.}]
bin/$(document_pdf): $(latex_code) bin/$(results)
	run container latex

bin/$(results): $(results_code)
	build image $(results_image)
	run container $(results_image)

clean:
	rm -rf bin/
\end{lstlisting}

The default target runs the container for the \LaTeX\ compilation while the \textit{results} target builds the image and runs the container that generates the \textit{results}.
When the \textit{author} edits a file from \textit{results code} or \textit{\LaTeX\ code}, only the required compilation is automatically triggered with Make.
Forced compilations without change could be done using the \textit{touch} utility that updates the modification time of a file.

The basic commands of the Makefile are the following:
\begin{lstlisting}[language=bash, style=lststyle, caption={Basic Makefile commands.}]
make                # Generate draft bin (fast).\ 
make FULL=1         # Generate full bin (slow).\ 
make clean          # Remove bin/ directory.
\end{lstlisting}

The \textit{author} may use the Makefile for developing \textit{documents} in the following way:
\begin{enumerate}
	\item clones/downloads the repository which consists of the Makefile and the set of files as depicted in Fig.\ref{fig:repositoryfiles},
	\item creates and develops the \textit{results code}.
	\item creates and develops the \textit{\LaTeX\ code}.
	\item executes \textit{make} for fast development iterations and experimentation,
	\item executes \textit{make FULL=1} for distribution or publication and
	\item executes \textit{make clean} to restore to a clean state.
\end{enumerate}

The only requirements that the Makefile imposes on \textit{code} is the presence of \textit{results code} that generates the draft and full versions of the \textit{document} and the presence of \textit{results code} and \textit{\LaTeX\ code} for saving/loading the \textit{results} to/from \textit{bin/} respectively.
An example of a \textit{results code} that generates the draft version \textit{document} in the field of neural networks would be a conditional that sets the number of epochs and training/validation/test samples to a low value.

\section{The Makefile utilities}
In this section we present useful utilities that can be combined with the Makefile to automate \textit{results} generation and improve the reproducibility of the \textit{document}.

A useful \LaTeX\ package for automatically presenting variables from \textit{results} in \textit{\LaTeX\ code} is \textit{datatool} with the following example use:
\begin{lstlisting}[language=TeX, style=lststyle, caption={\LaTeX\ datatool example of loading a file that contains pairs of keys and values (bin/keys-values.csv) generated by a \textit{results code} and getting the value of a key named lr.}]
\DTLloaddb{keys-values}{bin/keys-values.csv}
\DTLfetch{keys-values}{key}{lr}{value}
\end{lstlisting}

When using Python, a useful function for \textit{results code} is \textit{pandas.DataFrame.to\_latex} which automatically converts a dataframe table to a \LaTeX\ table:
\begin{lstlisting}[language=python, style=lststyle, caption={Convert Pandas DataFrame (df) to \LaTeX\ table (bin/metrics.tex) in \textit{results code}.}]
df = pd.DataFrame(table)
df.to_latex("bin/metrics.tex", float_format="%.2f")
\end{lstlisting}

There are some considerations need to be taken into account when reproducibility is needed.
Regarding \LaTeX\, a culprit against reproducibility is the time-date metadata embedded in the pdf output.
These can be disabled using the following commands into the \textit{\LaTeX\ code} or as extra arguments during \LaTeX\ compilation (as done by default in the Makefile):
\begin{lstlisting}[language=TeX, style=lststyle, caption={\LaTeX\ pdf reproducibility commands.}]
\pdfinfoomitdate=1
\pdfsuppressptexinfo=-1
\pdftrailerid{}
\end{lstlisting}

Regarding the \textit{results code} for Python, we can achieve deterministic stochasticity by setting the random seeds to a specific value such as:
\begin{lstlisting}[language=python, style=lststyle, caption={Python reproducibility commands for popular libraries.}]
# build-in random module
random.seed(0)
# numpy
np.random.seed(0)
# tensorflow
tf.random.set_seed(0)
# pytorch
torch.backends.cudnn.benchmark = False
torch.backends.cudnn.deterministic = True
torch.cuda.manual_seed_all(0)
torch.manual_seed(0)
\end{lstlisting}

Additionally, an \textit{author} can use the \textit{cmp} utility to compare \textit{documents} (generated in a different session, operating system or underlying hardware) byte by byte, and diffoscope for identifying sources of non-determinism as follows:
\begin{lstlisting}[language=bash, style=lststyle, caption={Test draft document version reproducibility. This can also be used as a test script when pushing or pull requesting to a remote repository.}]
# Generate the first draft document version.\ 
make
# Backup the first resulting pdf.\ 
cp bin/ms.pdf bin/ms-previous.pdf
# Update the modification timestamp of main.py.\ 
touch main.py
# Generate the second draft document version.\ 
make
# Compare draft document versions byte by byte.\ 
cmp bin/ms.pdf bin/ms-previous.pdf
# Compare draft document versions
# in a human readable form.\ 
diffoscope bin/ms.pdf bin/ms-previous.pdf
\end{lstlisting}

\section{Example use case of the Makefile}
This section provides a use case of the Makefile and also serves as a manual for developing \LaTeX\ \textit{documents} with Python as the \textit{results code} programming language.
We train, validate and test MobilenetV2 neural networks~\cite{sandler2018mobilenetv2} in PyTorch, on MNIST~\cite{lecun2010mnist}, FashionMNIST~\cite{xiao2017fashion}, KMNIST~\cite{clanuwat2018deep} and QMNIST~\cite{yadav2019cold}.
We use the default train, validation and test datasets and compare the use of ReLU, ReLU6\cite{dahl2013improving} and SiLU\cite{elfwing2018sigmoid} activation functions in the MobilenetV2 model.

@@

\section{Discussion}
The proposed Makefile is based on plain text file editing and can be used with any operating system that supports containers and Make.
Make has existed for decades and has passed the test of time, while the container technology was standardized with the Open Container Initiative and there exist alternatives such as Podman that could be used as drop-in replacements for Docker.
Moreover the Makefile can be combined with any version control system, container registry provider and text editor thus preventing `app/vendor lock-in' situations.

@@

\section*{Conclusion}
Creation, usage and publication of technical documents are one of the top challenges for reproducible technical documents~\cite{barba2019praxis}.
The proposed Makefile aids to this purpose by providing a way to write reproducible technical documents using the standard tools that many of the academics already use (\LaTeX, Docker, Make) in a portable, efficient and future-proof way.

>>>>

@@

\title{Principle of least action as a software development method}

@@

\begin{abstract}
	We propose principle of least action as a software development method.
	We release open source repositories of the examples use cases of this method.
\end{abstract}

\section{Introduction}
Action is ...
Principle of least action is ...

\section{Action and software development}
a. characteristics of the user (e.g. technical capabilities)
b. characteristics of the user environment (e.g. Operating system)
c. characteristics of the developer (e.g. technical capabilities)
choose the technologies that allow the developer to minimize the development and user path length through space (other computers) and time (in the future)
d. characteristics of the developer environment (e.g. Operating system)
different stakeholders calculate different actions. the developer sees the path that corresponds to the development procedure while the user sees the path that corresponds the usage of the binary
if the benefit is greater than the cost then it is implemented
moreover the cost and benefits are relative to the user and developer respectively
software development should take in consideration stability both for the user and the developer across two axis: space (other machines) and time
stability not only applies the same program use but also to other programs
the benefit of achieving a target is balanced with the cost of the time required to achieve it
the minimal action principle results in the unix philosophy: Programs should do one thing and one thing only. From that sense a window manager should only be responsible for managing windows and not workspaces, tags etc
the central concept of the paper is translating the idea of action and the principle of minimal action to software development as a specification and then show the implementation that uses Makefile+containers
present the design and implementation of fswm, then the Makefile for python and tex for document generation and then the Makefile for website creation
creating a program (the target) sometimes requires creating tools (as intermediate targets) that can be use to reduce the development time (the path length to the target)
it is not just about minimizing time or being completely accurate in satisfying the requirements. A combination of these is required
the path to the target is the source code which can be evaluated with make check. The path should fulfill the needs of the developers and can change
the target is the the final program which can be reached with make. The target should fulfill the needs of the users and can change
both users and developers want to reduce their relative action. Balance is achieved at a point
The principle of least action can be used throughout all decision of a software project.

\section{Development environment/tools}
Mention install-archlinux.
Mention vim etc.

\section{The Makefile}
The principle of least action has been applied in the following aspects of the Makefile:
\begin{itemize}
	\item this
	\item that
\end{itemize}

although we could have a target regarding downloading external data this makes the Makefile non-portable for other projects thus this action needs to be done within the code
consistency across time from the developer perspective means that the technology used to program it will be available in the future
consistency across time from the user perspective practically means that the program should be finished from the point of view of features and only bug fixes might be developed
makes use of pre-existing standards: GNU standard for Makefile conventions and POSIX for the use of UNIX tools
the portability of the Makefile is achieved by using the mandatory UNIX utilities also mentioned by the GNU standard with the exception of docker or npm where needed
there is an inverse relationship between consistency and configurability
think about that I have used subsets of standards
time includes user interation to reach the target and total development time (including maintainance)
users care about the program, developers care about the code
we define a target of reducing the time required to create and navigate windows in a computing environment
we use Makefile to create the isolated environment (container) by bootstraping all required files if not already present and to make sure that subsequent changes to the requirements update the environment. The DAG should be as project agnostic as possible thus sentinel files are used
we use these specific targets for the Makefile (all, check, clean, help) because the first three are standard targets recommended by GNU standard targets and the fourth one is used in the linux kernel

\section{Variable naming}
think prime-modifier-class naming scheme and how it relates to the action-as-information

\section{Example use case: window manager}
an extreme example of configurable window managers is `ricing' where the window manager is configured mostly for form and not function
fswm targets users with high technical capabilities (therefore no visual indication is needed regarding which window they are in every time etc.)
if keyboard portability is not so important then keycodes are hardcoded for the specific user environment
moreover a window manager should open windows only in full screen
the concept of tiling window managers is oriented towards multitasking
the key shortcuts of a window manager should be predictable by resembling as close as possible most other window managers and development environemnts
the need to be portable across different keyboards demands the use of an additional requirement (xcb-util-keysyms)
the set of functionalities that should be integrated to a window manager are: Enter next/previous window and open program in new window
the windows could be easily represented as a cyclic doubly linked list
thus C89 was chosen for optimal portability and xcb as the main library
an additional advantage of non-configurability is that the program is ready to be used when installed

\section{Example use case: technical document}
This section provides a use case of the Makefile and also serves as a manual for developing using this method.
We train, validate and test MobilenetV2 neural networks~\cite{sandler2018mobilenetv2} in PyTorch, on MNIST~\cite{lecun2010mnist}, FashionMNIST~\cite{xiao2017fashion}, KMNIST~\cite{clanuwat2018deep} and QMNIST~\cite{yadav2019cold}.
We use the default train, validation and test datasets and compare the use of ReLU, ReLU6\cite{dahl2013improving} and SiLU\cite{elfwing2018sigmoid} activation functions in the MobilenetV2 model.

@@

\section{About the use of Make}
The proposed Makefile is based on plain text file editing and can be used with any operating system that supports containers and Make.
Make has existed for decades and has passed the test of time, while the container technology was standardized with the Open Container Initiative and there exist alternatives such as Podman that could be used as drop-in replacements for Docker.
Moreover the Makefile can be combined with any version control system, container registry provider and text editor thus preventing `app/vendor lock-in' situations.

@@

\section*{Conclusion}
